---
title: "Chi Square Test of Independence for classes"
author: "kat"
date: "October 11, 2018"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
#gc()
#.rs.restartR()

```
### Purpose
- The purpose of this document is to expore relationships between the physics, chem, and bio classes
```{r}
rv.keep <- readRDS("rv.keepMASTER.rds")
#rv.keep
```
### Get 11th graders in 2017-2018
```{r}
full_11_2017 <- rv.keep[rv.keep$startYear == 2017,]
```

## #Biology Courses
- Bio courses
    + AP Biology, Biology, Biology 1, Biology 2, Honors Biology, Gen College Biology I w/Lab, Gen College Biology II w/Lab
```{r}
bio <- full_11_2017[ grepl("Biology" , full_11_2017$courseName ),] 
bioclasses <- unique(bio$courseName)
```

### only rows with bio classes
```{r}
full_11_2017_bio <- full_11_2017[full_11_2017$courseName %in% bioclasses,]
```

### Bio Grades Bar plot
```{r}
table_bio <- table(full_11_2017_bio$score)
barplot(table_bio, main = "grade distribution bio")
```

### Cleaning up Bio 
- Remove duplicated rows, spread & merge data
- Delete duplicate rows based on courseduplicator code 
- Convert data to wide
```{r}
bio_notdup <- full_11_2017_bio[!duplicated(full_11_2017_bio$courseduplicater),]
#take only personID,  scoreNum, courseIdentifier  ***numerical score
bio_short <- bio_notdup[,c(2, 50, 52)]

#Below is all 2017 juniors and their bio classes spread 
library(tidyr)
bio_wide <- spread(bio_short, courseIdentifier, scoreNum, fill=TRUE)
#bio_wide

```
## #create copies of bio columns just in case
```{r}
### CAREFUL WITH THIS!!!!!!!!!!!!! save as copy
bio_wideCopy <- bio_wide

bio_classes2 <- colnames(bio_wide[-c(1, 39,40, 41)])

# pass = 1, fail = 0
#cover all 1's to D's otherwise below wil not rewrite correctly
for (i in bio_classes2){
bio_wide[[i]] <- ifelse(bio_wide[[i]] == 4, 1,
                  ifelse(bio_wide[[i]] == 3, 1, 
                      ifelse(bio_wide[[i]] == 2, 1, 
                        ifelse(bio_wide[[i]] == 1, 0, 
                          ifelse(bio_wide[[i]] == "NA", "NA", 0)))))
}

#bio_wide
#bio_wideCopy

#convert semester classes differently
#- 4, 3, 2 = 2 credits
#- 1, 0 = 0 credits

semesterclasses <- c("XBIO111S1-Gen College Biology I w/Lab-S1", "XBIO111S2-Gen College Biology I w/Lab-S2", "XBIO112S2-Gen College Biology II w/Lab-S2")

#cover all 1's to D's otherwise below wil not rewrite correctly
for (i in semesterclasses){
bio_wide[[i]] <- ifelse(bio_wide[[i]] == 4, 2,
                  ifelse(bio_wide[[i]] == 3, 2, 
                      ifelse(bio_wide[[i]] == 2, 2, 
                        ifelse(bio_wide[[i]] == 1, 0, 
                          ifelse(bio_wide[[i]] == "NA", "NA", 0)))))
}
```

```{r}
#create row sum column
bio_wide$predSumBio <- rowSums(bio_wide[-1], na.rm=TRUE)
#bio_wide
```

### Bar plot of quarters 
```{r}
barplot(table(bio_wide$predSumBio),  main="Quarters passed of Biology, passed is C or greater")
barplot(prop.table(table(bio_wide$predSumBio)))
```

### BIO NUMBER OF QUARTERS PASSED
```{r}
bio_SUM <- bio_wide[c(1,42)]
#bio_SUM
```

### START CHEMISTRY, 2016 SOPHOMORES
####################### 

```{r}
full_10_2016 <- rv.keep[rv.keep$startYear == 2016,]

```

#Get Chemistry courses & Subset only Chem classes
```{r}
chem <- full_10_2016[ grepl("Chemistry" , full_10_2016$courseName ),] 
chemclasses <- unique(chem$courseName)
#chemclasses

full_10_2016_chem <- full_10_2016[full_10_2016$courseName %in% chemclasses,]
```

### Chem Grades Bar plot - ALL QUARTERS

```{r}
table_chem <- table(full_10_2016_chem$score)
barplot(table_chem)
```
### Cleaning up Chem
- Remove duplicated rows, spread & merge data
- Delete duplicate rows based on courseduplicator code 
- Convert data to wide
```{r}

chem_notdup <- full_10_2016_chem[!duplicated(full_10_2016_chem$courseduplicater),]
#take only personID,  scoreNum, courseIdentifier  ***numerical score
chem_short <- chem_notdup[,c(2, 50, 52)]
#chem_short

library(tidyr)
chem_wide <- spread(chem_short, courseIdentifier, scoreNum, fill=TRUE)
#chem_wide
```
### Recode Chem score values as Pass/Fail (2 or greater = pass)
```{r}
chem_wideCopy <- chem_wide
chem_classes2 <- colnames(chem_wide[-c(1)])
# pass = 1, fail = 0
#cover all 1's to D's otherwise below wil not rewrite correctly
for (i in chem_classes2){
chem_wide[[i]] <- ifelse(chem_wide[[i]] == 4, 1,
                  ifelse(chem_wide[[i]] == 3, 1, 
                      ifelse(chem_wide[[i]] == 2, 1, 
                        ifelse(chem_wide[[i]] == 1, 0, 
                          ifelse(chem_wide[[i]] == "NA", "NA", 0)))))
}

#chem_wide
#chem_wideCopy

```

### Create row sums
```{r}
#create row sum column
chem_wide$predSumChem <- rowSums(chem_wide[-1], na.rm=TRUE) #minus personID column
#chem_wide
```

```{r}
#create row sum column
bio_wide$predSumBio <- rowSums(bio_wide[-1], na.rm=TRUE)
#bio_wide

```

### Bar plot of quarters 
```{r}
barplot(table(chem_wide$predSumChem), main="Total Quarters Passed in Chem, passing = C")
barplot(prop.table(table(chem_wide$predSumChem)))
```

### CHEM NUMBER OF QUARTERS PASSED
```{r}
chem_SUM <- chem_wide[c(1,26)]
```

### START OF 2015 PHYSICS
##################################### 

### get 9th graders in 2015-2016 physivs
```{r}

full_9_2015 <- rv.keep[rv.keep$startYear == 2015,]
```


### Physics Courses

```{r}
physics <- full_9_2015[ grepl("Physics" , full_9_2015$courseName ),] 
physicsclasses<- unique(physics$courseName)
physicsclasses
```

### only rows with physics classes
```{r}
full_9_2015_phy <- full_9_2015[full_9_2015$courseName %in% physicsclasses,]
```
### Bio Grades Bar plot
```{r}
table_phys <- table(full_9_2015_phy$score)
barplot(table_phys, main="grade distribution physics")
```
### cleaning up Physics
- Remove duplicated rows, spread & merge data
- Delete duplicate rows based on courseduplicator code 
- Convert data to wide
```{r}
phys_notdup <- full_9_2015_phy[!duplicated(full_9_2015_phy$courseduplicater),]
#take only personID,  scoreNum, courseIdentifier  ***numerical score
phy_short <- phys_notdup[,c(2, 50, 52)]

#Below is all 2017 juniors and their bio classes spread 
library(tidyr)
phy_wide <- spread(phy_short, courseIdentifier, scoreNum, fill=TRUE)
#phy_wide

```

`

```{r}
#copy to compare scores
phy_wideCopy <- phy_wide

phy_classes2 <- colnames(phy_wide[-c(1)])

# pass = 1, fail = 0
#cover all 1's to D's otherwise below wil not rewrite correctly
for (i in phy_classes2){
phy_wide[[i]] <- ifelse(phy_wide[[i]] == 4, 1,
                  ifelse(phy_wide[[i]] == 3, 1, 
                      ifelse(phy_wide[[i]] == 2, 1, 
                        ifelse(phy_wide[[i]] == 1, 0, 
                          ifelse(phy_wide[[i]] == "NA", "NA", 0)))))
}

#phy_wide
#phy_wideCopy
```


### Create row sums
```{r}
#create row sum column
phy_wide$predSumPhy <- rowSums(phy_wide[-1], na.rm=TRUE) #minus personID column
#phy_wide
```

### Bar plot of quarters 
```{r}
barplot(table(phy_wide$predSumPhy), main="Total Quarters Passed in Physics, passing = C")
```


### PHYSICS NUMBER OF QUARTERS PASSED
```{r}
phy_SUM <- phy_wide[c(1,29)]
```

### COMBINING ALL THREE CLASSES
############# 

```{r}
#phy_SUM
#bio_SUM
#chem_SUM
one <- merge(phy_SUM, bio_SUM, by="personID", all=TRUE)
two <- merge(one, chem_SUM, by="personID", all=TRUE)
#two
two_numeric <- two
```

### ONLY 319 complete cases for now!!
```{r}
complete <- two[complete.cases(two),]
#complete
```

### Figure out how many non-NA entries  there are
- have trouble using complete cases for this section to remove them all so using above
```{r, eval=FALSE}
as.data.frame(sapply(two[-1], function(x) {
  factor(x, exclude=NULL) }))
two$predSumPhy <- factor(two$predSumPhy, exclude=NULL)
two$predSumBio <- factor(two$predSumBio, exclude=NULL)
two$predSumChem <- factor(two$predSumChem, exclude=NULL)
levels(two$predSumBio)
#summary(two)
```

### getting column means & medians

```{r}
library(robustbase)
#complete
colMeans(complete[-1])

colMedians(as.matrix(complete[-1]))

#summary(complete)
#str(complete)

```

###Ho: grade in physics is independent of grade in bio
###Ha: grade in physics is not independent of grade in bio

- Grades in Physics and Bio are not independent of each
```{r}
tbl <- table(complete$predSumPhy, complete$predSumBio)
chisq.test(tbl)
tbl
ctbl <- cbind(tbl[,"0"]+ tbl[,"1"], tbl[,"2"] , tbl[,"3"] ,tbl[,"4"])
ctbl <- rbind(ctbl["0",]+ ctbl["1",], ctbl["2",] ,ctbl["3",] , ctbl["4",])
ctbl
tbl
chisq.test(ctbl)
```

### PHysics & Bio WITH ALL VALUES NOT JUST complete cases


```{r}
tbl2 <- table(two_numeric$predSumPhy, two_numeric$predSumBio)
chisq.test(tbl2)
```


### Physics and Chemistry are not independent of each other
- More so than Phy & Bio
```{r}
tbl <- table(complete$predSumPhy, complete$predSumChem)
chisq.test(tbl)
tbl
ctbl <- cbind(tbl[,"0"]+ tbl[,"1"]+ tbl[,"2"] , tbl[,"3"] ,tbl[,"4"])
ctbl <- rbind(ctbl["0",]+ ctbl["1",]+ctbl["2",] ,ctbl["3",] , ctbl["4",])
sum(ctbl)
ctbl
tbl
chisq.test(ctbl)

```
### Chem & Bio
- Not independent of each other 

```{r}
tbl <- table(complete$predSumChem, complete$predSumBio)
chisq.test(tbl)
tbl
ctbl <- cbind(tbl[,"0"]+ tbl[,"1"]+ tbl[,"2"] , tbl[,"3"] ,tbl[,"4"])
ctbl <- rbind(ctbl["0",]+ ctbl["1",] + ctbl["2",] ,ctbl["3",] , ctbl["4",])
ctbl
sum(ctbl)
tbl
chisq.test(ctbl)

```

```{r}
chem.bio <- as.matrix(tbl)
rownames(chem.bio) <- paste('chem',0:4)
colnames(chem.bio) <- paste('bio',0:4)
chem.bio
```

```{r}
library(rcompanion)
results <- pairwiseNominalIndependence(chem.bio, fisher=FALSE, gtest=FALSE, chisq = TRUE, method="fdr")
results[results$p.adj.Chisq < .05,]
results[!results$p.adj.Chisq < .05,]
```

### THIS PLOTS ALL STUDENTS, NOT JUST COMPLETE CASES!!!

```{r}
phybioTable <- table(two$predSumPhy, two$predSumBio)
sum(phybioTable)
twocomplete <- two[complete.cases(two$predSumPhy & two$predSumBio),]
#How many students did better in Physics than Bio? 80
sum(twocomplete$predSumPhy > twocomplete$predSumBio, na.rm=TRUE)

# How many students did better in Bio than Physics 66
sum(twocomplete$predSumPhy < twocomplete$predSumBio, na.rm=TRUE)


phybioTable <- cbind(phybioTable[,"0"]+ phybioTable[,"1"] + phybioTable[,"2"] , phybioTable[,"3"] ,phybioTable[,"4"])
phybioTable <- rbind(phybioTable["0",]+ phybioTable["1",] + phybioTable["2",] ,phybioTable["3",] , phybioTable["4",])
phybioTable
```

```{r}
#reference https://ggplot2.tidyverse.org/reference/geom_count.html
# reference https://ggplot2.tidyverse.org/reference/geom_point.html
library(ggplot2)
ggplot(two, aes(predSumChem, predSumBio)) +
  geom_count(colour="gray")+
  geom_point(colour="gray")+
  geom_jitter(width=.2, height=.2) +
  geom_abline(intercept=0, slope=1, color="red", size=2)+
 geom_jitter(width=.2, height=.2, colour=ifelse((two$predSumBio==2 & two$predSumChem==4) |
                                                  (two$predSumBio==3 & two$predSumChem==4) |
                                                  (two$predSumBio==4 & two$predSumChem==3) |
                                                   (two$predSumBio==4 & two$predSumChem==4) , "blue", "gray"))


ggplot(two, aes(predSumChem, predSumPhy)) +
  geom_count(colour="gray")+
  geom_point(colour="gray") +
  geom_abline(intercept=0, slope=1, color="red", size=2)+
  geom_jitter(width=.2, height=.2, colour=ifelse((two$predSumPhy==4 & two$predSumChem==4) |
                                                  (two$predSumPhy==3 & two$predSumChem==4) |
                                                  (two$predSumPhy==4 & two$predSumChem==3) |
                                                   (two$predSumPhy==4 & two$predSumChem==2) , "blue", "gray")) 

ggplot(two, aes(predSumPhy, predSumBio)) +
  geom_point(colour="gray")+
  geom_count(colour="gray")+
  geom_abline(intercept=0, slope=1, color="red", size=2)+
  geom_jitter(width=.2, height=.2, colour=ifelse((two$predSumPhy==4 & two$predSumBio==4) |
                                                  (two$predSumPhy==3 & two$predSumBio==4) |
                                                  (two$predSumPhy==4 & two$predSumBio==3) , "blue", "gray")) 
```

### bar plot of grade distributions
```{r}

par(mfrow=c(2,2))
barplot(prop.table(table(chem_wide$predSumChem)), border="green")

barplot(prop.table(table(bio_wide$predSumBio)), col="red")
barplot(prop.table(table(phy_wide$predSumPhy)))
```